\PassOptionsToPackage{numbers,sort&compress}{natbib}
\documentclass{article}

\usepackage[final]{neurips_2025}

\makeatletter
\renewcommand{\@notice}{}
\makeatother

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{babel}

% \addto\captionsspanish{%
%   \renewcommand{\tablename}{Tabla}%
%   \renewcommand{\figurename}{Figura}%
% }

\usepackage{microtype}
\usepackage{amsmath,amssymb,amsfonts,bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{float}
\usepackage{placeins}

\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\textfraction}{0.07}
\renewcommand{\floatpagefraction}{0.8}

\captionsetup{justification=centering}

% Reduce espacio alrededor de figuras y tablas (sin romper NeurIPS)
\setlength{\textfloatsep}{10pt plus 2pt minus 2pt}
\setlength{\floatsep}{8pt plus 2pt minus 2pt}
\setlength{\intextsep}{10pt plus 2pt minus 2pt}

% Reduce espacio de captions
\setlength{\abovecaptionskip}{5pt}
\setlength{\belowcaptionskip}{0pt}

\title{Do World Models improve visual, continuous control tasks?}

\author{
José Andrés Ridruejo - Joaquín Mir - Javier Ríos \\
Final Project --- Deep Reinforcement Learning \\
}

\begin{document}

\maketitle

\begin{abstract}
    This paper investigates the impact of World Models on visual, continuous control tasks. We aim at answering the question of whether incorporating a World Model can enhance the performance of reinforcement learning agents in complex environments (in specific, we focus on the CarRacing-v3 environment). We compare the performance of agents with and without World Models, analyzing their learning efficiency, stability, and final performance. Our results suggest that World Models can significantly improve the agent's ability to learn and perform in visual, continuous control tasks, at the cost of increased computational complexity.
\end{abstract}

\section{Introduction}

Over the past months, we have been working on the main advantages that including Deep Learning models in the context of Reinforcement Learning can bring. In the first project, we focused on \textbf{representation learning} \cite{simclr} and how it can help to apply RL algorithms in complex environments (thus tackling the curse of dimensionality). The second project revolved around \textbf{model-free discrete control}, where we used implementations of Rainbow DQN and its variants \cite{rainbow2018combining}, in which agents learn the \textit{Q-functions} directly from visual observations. Finally, the third project extended this framework to \textbf{continuous control} tasks, using Actor-Critic methods, mainly PPO \cite{schulman2017proximal} and SAC \cite{haarnoja2018soft}, discretizing continuous action spaces and analyzing on-policy vs. off-policy trade-offs from pixel-based observations.

While these approaches demonstrated effective learning of locomotion policies in \textit{MuJoCo} environments, they all share a fundamental limitation: \textbf{sample inefficiency}. Each gradient update relies on data collected from interactions with the environment, which becomse prohibitive in complex environments or physically constrained settings like robotics. This motivates our final project: transitioning to a \textbf{model-based} paradigm, where we can leverage a learned model of the environment to generate synthetic data and improve learning efficiency.

This final project builds on the previous ones, as we will be using similar environments and algorithms, but now we will be incorporating \textbf{World Models} \cite{ha2018world}, which are generative models that learn to predict future states of the environment. Instead of learning a policy directly from the environment, we will first learn a World Model that captures the dynamics of the environment, and then use it to train a policy. This approach has the potential to improve sample efficiency and generalization, as the agent can learn from imagined trajectories generated by the World Model.

Assuming a well-trained World Model, the agent (f.e. an actor-critic algorithm) can operate solely in the compact latent space rather than the high-dimensional pixel space. This can lead to faster learning and better generalization, as the agent can focus on the underlying dynamics of the environment rather than the raw visual input. Additionally, the World Model can be used for planning and decision-making, allowing the agent to anticipate future states and make informed choices based on predicted outcomes.
\newpage
We implement this using \textbf{R2-Dreamer} \cite{morihira2026r2dreamer}, an ICLR 2026 state-of-the-art evolution of DreamerV3 \cite{hafner2023mastering} that eliminates costly image reconstructions in favor of redundancy reduction in the latent space.

The core research question is clear: \textbf{can a world model outperform our model-free baselines in visual, continuous control tasks? And, if so, how does it affect sample efficiency and final performance?} We will evaluate this question in order to assess whether the added complexity of appending a World Model is justified by the performance gains it provides (which, we hypothesize, will depend on the specific environment and task at hand).

% \paragraph{Contribuciones.}
% \begin{itemize}
%     \item Proponemos un enfoque de detección de fraude basado en incertidumbre que integra predicción y decisión operativa.
%     \item Comparamos una BNN con baselines deterministas, evaluando tanto rendimiento predictivo como utilidad decisional.
%     \item Introducimos una política de triaje (\textsc{Accept/Review/Block}) basada en probabilidad e incertidumbre, junto con análisis de predicción selectiva.
%     \item Mostramos que el umbral de incertidumbre no es un hiperparámetro fijo arbitrario, sino una palanca operativa que puede ajustarse para equilibrar automatización y riesgo residual.
% \end{itemize}

\subsection{Related Work}

\subsubsection{Model-Free Baselines}

Our model-free baselines (Rainbow DQN and PPO/SAC) are well-established algorithms in the field of reinforcement learning, known for their effectiveness in various environments. Rainbow DQN combines six improvements to the original DQN algorithm (double Q-learning, prioritized experience replay, dueling networks, multi-step learning, distributional RL, and noisy nets)  for discrete action spaces, achieving strong performance in Atari games \cite{rainbow2018combining}, but struggling with continuous control tasks

PPO and SAC, on the other hand, are able to handle continuous actions implicitly (by means of policy gradients and maximum entropy RL) but remain sample-inefficient as each update consumes fresh rollouts without data reuse.

\subsubsection{World Models}

World models learn \textbf{generative dynamic models} to enable \textit{imagination-based} policy optimization \cite{ha2018world}. They are based (as many other ideas in the field) on the way humans learn and interact with the world:

\begin{quote}
    Humans develop a mental model of the world based on what they are able to perceive with their limited senses. The decisions and actions we make are based on this internal model \,\dots\, We are able to observe a scene and remember an abstract description thereof.
\end{quote}

Early approaches to this new paradigm, such as PlaNet \cite{hafner2019planet}, used deterministic RNNs for latent prediction, but suffered from compounding errors and limited performance. DreamerV3 \cite{hafner2023mastering} (developed after 4 years by the same team) introduced some key innovations, such as the use of a \textbf{recurrent state-space model} (RSSM) with stochastic latents, and the use of \textbf{latent imagination} for policy optimization, which significantly improved performance and sample efficiency. Despite its success in DMC, Atari and Crafter environments, the decoder CNN used for image reconstruction was a major bottleneck in terms of computational cost and scalability.

R2-Dreamer \cite{morihira2026r2dreamer} addresses this issue by eliminating the need for image reconstruction. Instead, it relies on \textbf{Barlow Twins-style redundancy reduction} \cite{zbontar2021barlow}, which implements an objective function that aims at making the cross-correlation matrix between two views of the latent space as close to the identity as possible. Mathematically, BT enforces:
\begin{equation}
    \mathcal{L}_{BT} = \sum_{i} (1 - C_{ii})^2 + \lambda \sum_{i} \sum_{j \neq i} C_{ij}^2
\end{equation}
where $C$ is the cross-correlation matrix between two views of the latent space, and $\lambda$ is a hyperparameter that controls the trade-off between invariance and redundancy reduction. This produces compact, decorrelated latents without the need for pixel-level supervision, thus enabling a \textbf{1.6$\times$ training speedup} over DreamerV3 while maintaining or improving performance across a wide range of tasks.

More recently, other approaches such as Next Embedding-Dreamer \cite{nedreamer2026} and ResWM \cite{reswm2026} have further improved the performance of world models by introducing new architectures and training techniques, such as contrastive learning and residual connections. In spite of this, R2-Dreamer remains the candidate par excellence for our project, as it strikes a good balance between performance and computational cost.

% Table~\ref{tab:sota} positions our baselines against 2026 SOTA on DMC locomotion (normalized scores at 1M steps, 12M params).

% \begin{table}[ht]
%     \centering
%     \caption{SOTA comparison on DMC locomotion (2026). R2-Dreamer leads efficiency.}
%     \label{tab:sota}
%     \begin{tabular}{lcc}
%         \toprule
%         Method              & DMC Score        & Speedup vs DreamerV3 \\
%         \midrule
%         Rainbow DQN         & 50--70           & --                   \\
%         PPO/SAC             & 70--90           & --                   \\
%         DreamerV3           & 95--100          & 1.0$\times$          \\
%         \textbf{R2-Dreamer} & \textbf{98--102} & \textbf{1.6$\times$} \\
%         ResWM               & 105--110         & 1.2$\times$          \\
%         \bottomrule
%     \end{tabular}
% \end{table}

\section{Method}

\section{Results}

\section{Discussion}

\section{Conclusion}

\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
